Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-12 ind_files/99_Appendix_D_examples.tex
Add a figure comparing two realizations of stoch logistic eqn

\section{An example in one dimension: The stochastic logistic equation}
Here, we analyze example \ref{ex_1D_stoch_logistic}. To recap, we had a population of individuals that exhibit a constant per-capita birth rate $\lambda > 0$, and a per-capita death rate that had the linear density-dependence $d(n) = \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n$, where $\mu$ and $K$ are positive constants. Thus, we have the equations
\begin{equation}
\label{ex_1D_stoch_logistic_BD_eqns}
\begin{aligned}
    b(n) &= \lambda n\\
    d(n) &= \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n
\end{aligned}
\end{equation}
Here, $K$ is the system-size parameter. Introducing the new variable $x=n/K$, we obtain
\begin{align*}
    b_K(x) &= \frac{1}{K}b(n) = \frac{1}{K}\lambda Kx\\
    d_K(x) &= \frac{1}{K}d(n) = \frac{1}{K}\left(\mu + (\lambda-\mu)\frac{Kx}{K}\right)Kx
\end{align*}
Thus, we have
\begin{equation*}
    A^{\pm}(x) = b_K(x)\pm d_K(x) = x\left(\lambda \pm \left(\left(\mu + (\lambda-\mu)x\right)\right) \right)
\end{equation*}
Defining $r=\lambda-\mu$ and $v=\lambda+\mu$ and using equation \eqref{1D_SDE}, we 
see that the `mesoscopic view' of the system is given by the solution of the SDE
\begin{equation}\label{ex_1D_stoch_logistic_full_SDE}
dX_t =  rX_t(1-X_t)dt + \sqrt{\frac{X_t(v+rX_t)}{K}}dB_t
\end{equation}
From equation \eqref{1D_det_limit}, we see that the deterministic dynamics are
\begin{equation}\label{ex_1D_stoch_logistic_det_limit}
\frac{dx}{dt} = A^-(x) = rx(1-x)
\end{equation}
showing that in the infinite population limit, we
obtain the logistic equation. Letting $\alpha(t)$ be the solution of the logistic equation \eqref{ex_1D_stoch_logistic_det_limit}, We can Taylor expand $A^{\pm}(x)$ for the weak noise approximation, and we find:
\begin{align*}
A^-_1(x) &= \frac{d}{dx}(rx(1-x))\biggl{|}_{x=\alpha} = r(1 - 2\alpha(t))\\
A^+_0(x) &= \alpha(t)(v+r\alpha(t))
\end{align*}
Thus, the weak noise approximation of \ref{ex_1D_stoch_logistic_BD_eqns} is given by
\begin{equation}
    X_t = \alpha(t) + \frac{1}{\sqrt{K}}Y_t
\end{equation}
where the stochastic process $Y_t$ is an Ornstein-Uhlenbeck process given by the solution to the linear SDE
\begin{align}
    dY_t &= A^-_1(t)Y_tdt + \sqrt{A^+_0(t)}dB_t\nonumber\\
    \Rightarrow dY_t &= r(1 - 2\alpha(t))Y_tdt + \sqrt{\alpha(t)(v+r\alpha(t))}dB_t\label{ex_1D_stoch_logistic_WNA}
\end{align}

@@

The deterministic trajectory \eqref{ex_1D_stoch_logistic_det_limit} has two fixed points, one at $x=0$ (extinction) and one at $x=1$ (corresponding to a population size of $n=K$). For $r > 0$, $x=0$ is unstable and $x=1$ is a global attractor, meaning in the deterministic limit, when $r > 0$, all populations end up at $x=1$ given enough time. The stochastic dynamics \eqref{ex_1D_stoch_logistic_full_SDE} and \eqref{ex_1D_stoch_logistic_WNA} depend not only on $r$, but also on $v$, the sum of the birth and death rates. It has been proven that $X_t = 0$ is the only recurrent state for the full stochastic dynamics \eqref{ex_1D_stoch_logistic_full_SDE}, meaning that every population is guaranteed to go extinct\footnote{This can be proven using tools from Markov chain theory. For those interested, the proof uses ergodicity to arrive at a contradiction if any state other than $0$ exhibits a non-zero density at steady state.} given enough time~\citep{nasell_extinction_2001}, thus illustrating an important difference between finite and infinite populations. $X_t = 0$ is also an `absorbing' state since once a population goes extinct, it has no way of being revived in this model. However, if $K$ is large enough, the eventual extinction of the population may take a very long time. In fact, we can make the expected time to extinction arbitrarily long by making $K$ sufficiently large. Thus, for moderately large values of $K$, it is biologically meaningful only to look at a weaker version of the steady state distribution by imposing the condition that the population does not go extinct and looking at the `transient' dynamics~\citep{hastings_transients_2004}. Conditioned on non-extinction, the solution to \eqref{ex_1D_stoch_logistic_full_SDE} has a `quasistationary' distribution about the deterministic attractor $X_t = 1$, with some variance reflecting the effect of noise-induced fluctuations in population size~\citep{nasell_extinction_2001} due to the finite size of the population. The weak-noise approximation \eqref{ex_1D_stoch_logistic_WNA} implicitly assumes non-extinction by only measuring small fluctuations from the deterministic solution to \eqref{ex_1D_stoch_logistic_det_limit} and thus, at steady state, naturally describes a quasistationary distribution centered about $X_t = 1$. The steady-state density (probability density function as $t \to \infty$) of the exact birth-death process \eqref{ex_1D_stoch_logistic_BD_eqns} is compared with that predicted by \eqref{ex_1D_stoch_logistic_full_SDE} and \eqref{ex_1D_stoch_logistic_WNA} for various values of $K$ in figure \ref{fig_1D_stoch_logistic_densities}.

\myfig{0.85}{Media/4.2_stoch_logistic_distributions.png}{Comparison of the steady-state densities given by \eqref{ex_1D_stoch_logistic_BD_eqns}, \eqref{ex_1D_stoch_logistic_full_SDE}, and \eqref{ex_1D_stoch_logistic_WNA} for \textbf{(A)} $K = 500$, \textbf{(B)} $K = 1000$, and \textbf{(C)} $K = 10000$. $\lambda = 2, \mu = 1$ for all thee cases. Each curve was obtained using $1000$ independent realizations.}{fig_1D_stoch_logistic_densities}

@@

Going from population numbers $\mathbf{v}$ to densities $\mathbf{x} = \mathbf{v}/K$, we obtain the birth and death rates:
\begin{equation}
\label{nD_example_density_b_d_rates}
\begin{aligned}
b^{(K)}_i(\mathbf{x}) &= \left(\lambda + \sum\limits_{j=1}^{m}\beta_{ij}x_j\right)x_i\\
d^{(K)}_i(\mathbf{x}) &= \left(\mu + \sum\limits_{j=1}^{m}\delta_{ij}x_j\right)x_i
\end{aligned}
\end{equation}
Thus, we have
\begin{equation*}
A^{\pm}_{i} = x_i\left(\left(\lambda \pm \mu\right) + \sum\limits_{j=1}^{m}\left(\beta_{ij} \pm \delta_{ij}\right)x_j\right)
\end{equation*}
Defining $r \coloneqq \lambda - \mu$, $\nu \coloneqq \lambda + \mu$, and $T_{ij} \coloneqq \beta_{ij} + \delta_{ij}$, we see from equation \eqref{nD_Ito_SDE} that the mesoscopic view is the $m$ dimensional SDE given by
\begin{equation}
\label{nD_example_SDE}
d\mathbf{X}_{t} = \mathbf{A^-}(\mathbf{X}_t)dt + \frac{1}{\sqrt{K}}\mathbf{D}(\mathbf{X}_t)d\mathbf{B}_t
\end{equation}
where 
\begin{equation*}
\mathbf{A^-}_i = {(\mathbf{X}_{t})}_i(r + \sum\limits_{j=1}^{m}M_{ij}{(\mathbf{X}_{t})}_j) 
\end{equation*}
and
\begin{equation*}
(\mathbf{D}\mathbf{D}^{\mathrm{T}})_i = {(\mathbf{X}_{t})}_i(\nu + \sum\limits_{j=1}^{m}T_{ij}{(\mathbf{X}_{t})}_j) 
\end{equation*}
From \eqref{nD_det_limit}, we see that the deterministic limit is a set of $m$ coupled ODEs given by
\begin{equation}
\label{nD_example_det_limit}
\frac{d x_i}{dt} = x_i\left(r + \sum\limits_{j=1}^{m}M_{ij}x_j\right)
\end{equation}
These are precisely the Lotka-Volterra equations for a system of $m$ species. By matching the terms of \eqref{nD_example_density_b_d_rates} with those of \eqref{nD_functional_forms_for_replicator}, we can identify that we have $\mu = 0$ and
\begin{equation}
\label{nD_example_fitness}
w_i(\mathbf{x}) = r + \sum\limits_{j=1}^{m}M_{ij}x_j
\end{equation}
If $p_i(t)$ is the frequency of type $i$ individuals in the population at time $t$ and $N_K(t) = \sum_i x_i(t)$, then the mean fitness is given by
\begin{align}
\overline{w}(t) &= \sum\limits_{i=1}^{m}w_ip_i\\
&= \sum\limits_{i=1}^{m}\left(r + \sum\limits_{j=1}^{m}M_{ij}x_j\right)p_i\\
&= r + \sum\limits_{i=1}^{m}p_i\left(\sum\limits_{j=1}^{m}M_{ij}x_j\right)
\end{align}
where we have used the fact that $\sum_i p_i = 1$ in the last line. Using \eqref{nD_replicator_mutator} to write down the equations for the frequencies $p_i$, we obtain
\begin{equation}
\frac{1}{N_K(t)}\frac{dp_i}{dt} = \left[(\mathbf{Mp})_i - \mathbf{p}\cdot\mathbf{Mp}\right]p_i   
\end{equation}
which is the familiar version of the replicator equation seen in most textbooks, with an extra $N_K(t)$ factor to account for the fact that $\sum_i x_i$ is allowed to fluctuate in our model. If instead $N_K$ was a constant for all time, it could simply be absorbed into the definition of the payoff matrix $M$ to obtain exactly the replicator equation as presented in most ecology/evolution textbooks. Both the stochastic dynamics \eqref{nD_example_SDE} and the deterministic limit \eqref{nD_example_det_limit} can be simplified from an $m$ dimensional system to an $m-1$ dimensional system by a coordinate transformation that projects the dynamics onto an appropriate curve: If we go from the variables $x_1,\ldots,x_m$ to the variables $p_1,\ldots,p_{m-1},N_K$, we can exploit the fact that $N_K$ varies much less than the $p_i$ terms to project the system onto a `slow manifold' in which $N_K$ is approximately constant, thus obtaining an $m-1$ dimensional system of equations and recovering the relation between the Lotka-Volterra equations for $m$ species and the replicator equation for $m-1$ tactics~\citep{constable_mapping_2017,parsons_dimension_2017}. However, I will not explore such dimensional reduction techniques further in this manuscript, and refer the reader to~\citep{constable_stochastic_2013} and~\citep{parsons_dimension_2017} for a review of the ideas of (stochastic) dynamics on slow manifolds.

@@

We can also calculate $\mathcal{D}_{\zeta}[\mathcal{A}^-]$ as
\begin{align*}
\mathcal{D}_{\zeta}[\mathcal{A}^-] &= \frac{d}{d\epsilon}\left( r\int\limits_{\mathcal{T}} m(x,y)(\psi(y)+\epsilon\zeta(y))dy - \frac{\psi(x)+\epsilon\zeta(x)}{n(x)}\int\limits_{\mathcal{T}}\alpha(x,y)(\psi(y)+\epsilon\zeta(y))dy\right) \biggl{|}_{\epsilon = 0}\\
&= r\int\limits_{\mathcal{T}}m(x,y)\zeta(y)dy - \frac{1}{n(x)}\left(\psi(x)\int\limits_{\mathcal{T}}\alpha(x,y)\zeta(y)dy + \zeta(x)\int\limits_{\mathcal{T}}\alpha(x,y)\psi(y)dy\right)
\end{align*}
Using this in equation \eqref{spatial_covariance_zeta},
\cite{rogers_demographic_2012} (and later~\cite{rogers_modes_2015}) have shown that the contribution of demographic stochasticity can lead to inhibition of branching, and thus, while the population undergoes infinitely many branching events in the infinite population prediction, this does not happen for finite populations.
\myfig{0.8}{Media/2.1_dem_stoch_effects.png}{\textbf{Effect of population size on evolutionary branching}. Two different realizations of the system \eqref{Rogers_logistic_BD} with $n(x) = \exp((x^2)/\sigma_{K}^2)$. Simulation parameters are $\sigma_{K} = 1.9, \sigma_{\alpha} = 0.7, \sigma^2_{m} = 0.05$ for \textbf{top:} $K = 1000$ and \textbf{bottom:} $K = 10000$. Each point represents an individual. Note that the model on top remains monomorphic whereas the model on the bottom exhibits evolutionary branching, where an initially monomorphic population evolves to become dimorphic.}{fig_dem_stoch}
It is left as an exercise for the reader to verify by the same steps that if we instead have the birth rate functional $b(x|\phi) = \lambda\int m(x,y)\phi(y)dy$ (with $m(x,y)$ as defined in \eqref{Rogers_logistic_BD}) and the death rate functional $d(x|\phi) = \phi(x)\left(\mu+(\lambda-\mu)\phi(x)/K\right)$, the infinite-population limit yields the famous Fisher-KPP equation with growth rate $r=\lambda-\mu$ and diffusion constant $D = \lambda \sigma_m^2/2$.

>>>>

\section{An example in one dimension: The stochastic logistic equation}
Here, we analyze example \ref{ex_1D_stoch_logistic}. To recap, we had a population of individuals that exhibit a constant per-capita birth rate $\lambda > 0$, and a per-capita death rate that had the linear density-dependence $d(n) = \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n$, where $\mu$ and $K$ are positive constants. Thus, we have the equations
\begin{equation}
\label{ex_1D_stoch_logistic_BD_eqns}
\begin{aligned}
    b(n) &= \lambda n\\
    d(n) &= \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n
\end{aligned}
\end{equation}
Here, $K$ is the system-size parameter. Introducing the new variable $x=n/K$, we obtain
\begin{align*}
    b_K(x) &= \frac{1}{K}b(n) = \frac{1}{K}\lambda Kx\\
    d_K(x) &= \frac{1}{K}d(n) = \frac{1}{K}\left(\mu + (\lambda-\mu)\frac{Kx}{K}\right)Kx
\end{align*}
Thus, we have
\begin{equation*}
    A^{\pm}(x) = b_K(x)\pm d_K(x) = x\left(\lambda \pm \left(\left(\mu + (\lambda-\mu)x\right)\right) \right)
\end{equation*}
Defining $r=\lambda-\mu$ and $v=\lambda+\mu$ and using equation \eqref{1D_SDE}, we 
see that the `mesoscopic view' of the system is given by the solution of the SDE
\begin{equation}\label{ex_1D_stoch_logistic_full_SDE}
dX_t =  rX_t(1-X_t)dt + \sqrt{\frac{X_t(v+rX_t)}{K}}dB_t
\end{equation}
From equation \eqref{1D_det_limit}, we see that the deterministic dynamics are
\begin{equation}\label{ex_1D_stoch_logistic_det_limit}
\frac{dx}{dt} = A^-(x) = rx(1-x)
\end{equation}
showing that in the infinite population limit, we obtain the logistic equation. This derivation also makes it clear that two systems with very different stochastic dynamics can nevertheless converge to the same infinite population limit \eqref{ex_1D_stoch_logistic_det_limit}, since \eqref{ex_1D_stoch_logistic_det_limit}t only depends on the difference $\lambda-\mu$.
\myfig{0.6}{Media/App_stoch_logistic_same_w_different_t.png}{Comparison of time series (Gillespie simulations) for two different parameter values of the stochastic logistic equation. Simulations with the parameter values $\lambda = 4, \mu = 2$ are plotted in red, and simulations with the parameter values $\lambda = 31, \mu = 29$ are plotted in blue. Both simulations have $K=200$. The infinite population limit of both simulations is plotted in black dotted lines. The graph shows 10 realizations each for the two sets of parameter values.}{fig_1D_same_w_different_t_comparison}
To illustrate the effects of this seemingly innocent fact, figure \ref{fig_1D_same_w_different_t_comparison} compares two simulations which have the same value of $\lambda - \mu$ but a ten-fold difference in $\lambda+\mu$. As is clear from the figure, though both populations have the same behavior at the infinite population limit, populations with a higher value of $\lambda+\mu$ exhibit much wilder fluctuations and are therefore more prone to stochastic extinction. This is the root cause of the noise-induced selection that occurs in higher dimensions, discussed in detail in part \ref{part_summary}.

Letting $\alpha(t)$ be the solution of the logistic equation \eqref{ex_1D_stoch_logistic_det_limit}, We can Taylor expand $A^{\pm}(x)$ for the weak noise approximation, and we find:
\begin{align*}
A^-_1(x) &= \frac{d}{dx}(rx(1-x))\biggl{|}_{x=\alpha} = r(1 - 2\alpha(t))\\
A^+_0(x) &= \alpha(t)(v+r\alpha(t))
\end{align*}
Thus, the weak noise approximation of \ref{ex_1D_stoch_logistic_BD_eqns} is given by
    X_t = \alpha(t) + \frac{1}{\sqrt{K}}Y_t
where the stochastic process $Y_t$ is an Ornstein-Uhlenbeck process given by the solution to the linear SDE
    dY_t &= A^-_1(t)Y_tdt + \sqrt{A^+_0(t)}dB_t\nonumber\\
    \Rightarrow dY_t &= r(1 - 2\alpha(t))Y_tdt + \sqrt{\alpha(t)(v+r\alpha(t))}dB_t\label{ex_1D_stoch_logistic_WNA}

@@

The deterministic trajectory \eqref{ex_1D_stoch_logistic_det_limit} has two fixed points, one at $x=0$ (extinction) and one at $x=1$ (corresponding to a population size of $n=K$). For $r > 0$, $x=0$ is unstable and $x=1$ is a global attractor, meaning in the deterministic limit, when $r > 0$, all populations end up at $x=1$ given enough time.\myfig{0.85}{Media/4.2_stoch_logistic_distributions.png}{Comparison of the steady-state densities given by \eqref{ex_1D_stoch_logistic_BD_eqns}, \eqref{ex_1D_stoch_logistic_full_SDE}, and \eqref{ex_1D_stoch_logistic_WNA} for \textbf{(A)} $K = 500$, \textbf{(B)} $K = 1000$, and \textbf{(C)} $K = 10000$. $\lambda = 2, \mu = 1$ for all thee cases. Each curve was obtained using $1000$ independent realizations.}{fig_1D_stoch_logistic_densities}
The stochastic dynamics \eqref{ex_1D_stoch_logistic_full_SDE} and \eqref{ex_1D_stoch_logistic_WNA}, however, depend not only on $r$, but also on $v$, the sum of the birth and death rates. It has been proven that $X_t = 0$ is the only recurrent state for the full stochastic dynamics \eqref{ex_1D_stoch_logistic_full_SDE}, meaning that every population is guaranteed to go extinct\footnote{This can be proven using tools from Markov chain theory. For those interested, the proof uses ergodicity to arrive at a contradiction if any state other than $0$ exhibits a non-zero density at steady state.} given enough time~\citep{nasell_extinction_2001}, thus illustrating an important difference between finite and infinite populations. $X_t = 0$ is also an `absorbing' state since once a population goes extinct, it has no way of being revived in this model. However, if $K$ is large enough, the eventual extinction of the population may take a very long time. In fact, we can make the expected time to extinction arbitrarily long by making $K$ sufficiently large. Thus, for moderately large values of $K$, it is biologically meaningful only to look at a weaker version of the steady state distribution by imposing the condition that the population does not go extinct and looking at the `transient' dynamics~\citep{hastings_transients_2004}.
Conditioned on non-extinction, the solution to \eqref{ex_1D_stoch_logistic_full_SDE} has a `quasistationary' distribution about the deterministic attractor $X_t = 1$, with some variance reflecting the effect of noise-induced fluctuations in population size~\citep{nasell_extinction_2001} due to the finite size of the population. The weak-noise approximation \eqref{ex_1D_stoch_logistic_WNA} implicitly assumes non-extinction by only measuring small fluctuations from the deterministic solution to \eqref{ex_1D_stoch_logistic_det_limit} and thus, at steady state, naturally describes a quasistationary distribution centered about $X_t = 1$. The steady-state density (probability density function as $t \to \infty$) of the exact birth-death process \eqref{ex_1D_stoch_logistic_BD_eqns} is compared with that predicted by \eqref{ex_1D_stoch_logistic_full_SDE} and \eqref{ex_1D_stoch_logistic_WNA} for various values of $K$ in figure \ref{fig_1D_stoch_logistic_densities}.

@@

Going from population numbers $\mathbf{v}$ to densities $\mathbf{x} = \mathbf{v}/K$, we obtain the birth and death rates:
\begin{equation}
\label{nD_example_density_b_d_rates}
\begin{aligned}
b^{(K)}_i(\mathbf{x}) &= \left(\lambda + \sum\limits_{j=1}^{m}\beta_{ij}x_j\right)x_i\\
d^{(K)}_i(\mathbf{x}) &= \left(\mu + \sum\limits_{j=1}^{m}\delta_{ij}x_j\right)x_i
\end{aligned}
\end{equation}
Thus, we have
\begin{equation*}
A^{\pm}_{i} = x_i\left(\left(\lambda \pm \mu\right) + \sum\limits_{j=1}^{m}\left(\beta_{ij} \pm \delta_{ij}\right)x_j\right)
\end{equation*}
Defining $r \coloneqq \lambda - \mu$, $\nu \coloneqq \lambda + \mu$, and $T_{ij} \coloneqq \beta_{ij} + \delta_{ij}$, we see from equation \eqref{nD_Ito_SDE} that the mesoscopic view is the $m$ dimensional SDE given by
\begin{equation}
\label{nD_example_SDE}
d\mathbf{X}_{t} = \mathbf{A^-}(\mathbf{X}_t)dt + \frac{1}{\sqrt{K}}\mathbf{D}(\mathbf{X}_t)d\mathbf{B}_t
\end{equation}
where 
\begin{equation*}
\mathbf{A^-}_i = {(\mathbf{X}_{t})}_i(r + \sum\limits_{j=1}^{m}M_{ij}{(\mathbf{X}_{t})}_j) 
\end{equation*}
and
\begin{equation*}
(\mathbf{D}\mathbf{D}^{\mathrm{T}})_i = {(\mathbf{X}_{t})}_i(\nu + \sum\limits_{j=1}^{m}T_{ij}{(\mathbf{X}_{t})}_j) 
\end{equation*}
From \eqref{nD_det_limit}, we see that the deterministic limit is a set of $m$ coupled ODEs given by
\begin{equation}
\label{nD_example_det_limit}
\frac{d x_i}{dt} = x_i\left(r + \sum\limits_{j=1}^{m}M_{ij}x_j\right)
\end{equation}
These are precisely the Lotka-Volterra equations for a system of $m$ species. By matching the terms of \eqref{nD_example_density_b_d_rates} with those of \eqref{nD_functional_forms_for_replicator}, we can identify that we have $\mu = 0$ and
\begin{equation}
\label{nD_example_fitness}
w_i(\mathbf{x}) = r + \sum\limits_{j=1}^{m}M_{ij}x_j
\end{equation}
If $p_i(t)$ is the frequency of type $i$ individuals in the population at time $t$ and $N_K(t) = \sum_i x_i(t)$, then the mean fitness is given by
\begin{align}
\overline{w}(t) &= \sum\limits_{i=1}^{m}w_ip_i\\
&= \sum\limits_{i=1}^{m}\left(r + \sum\limits_{j=1}^{m}M_{ij}x_j\right)p_i\\
&= r + \sum\limits_{i=1}^{m}p_i\left(\sum\limits_{j=1}^{m}M_{ij}x_j\right)
\end{align}
where we have used the fact that $\sum_i p_i = 1$ in the last line. Using \eqref{nD_replicator_mutator} to write down the equations for the frequencies $p_i$, we obtain
\begin{equation}
\frac{1}{N_K(t)}\frac{dp_i}{dt} = \left[(\mathbf{Mp})_i - \mathbf{p}\cdot\mathbf{Mp}\right]p_i   
\end{equation}
which is the familiar version of the replicator equation seen in most textbooks, with an extra $N_K(t)$ factor to account for the fact that $\sum_i x_i$ is allowed to fluctuate in our model. If instead $N_K$ was a constant for all time, it could simply be absorbed into the definition of the payoff matrix $M$ to obtain exactly the replicator equation as presented in most ecology/evolution textbooks. Both the stochastic dynamics \eqref{nD_example_SDE} and the deterministic limit \eqref{nD_example_det_limit} can be simplified from an $m$ dimensional system to an $m-1$ dimensional system by a coordinate transformation that projects the dynamics onto an appropriate curve: If we go from the variables $x_1,\ldots,x_m$ to the variables $p_1,\ldots,p_{m-1},N_K$, we can exploit the fact that $N_K$ varies much less than the $p_i$ terms to project the system onto a `slow manifold' in which $N_K$ is approximately constant, thus obtaining an $m-1$ dimensional system of equations and recovering the relation between the Lotka-Volterra equations for $m$ species and the replicator equation for $m-1$ tactics~\citep{constable_mapping_2017,parsons_dimension_2017}. However, I will not explore such dimensional reduction techniques further in this manuscript, and refer the reader to~\cite{constable_stochastic_2013} and~\cite{parsons_dimension_2017} for a review of the ideas of (stochastic) dynamics on slow manifolds.

@@

We can also calculate $\mathcal{D}_{\zeta}[\mathcal{A}^-]$ as
\begin{align*}
\mathcal{D}_{\zeta}[\mathcal{A}^-] &= \frac{d}{d\epsilon}\left( r\int\limits_{\mathcal{T}} m(x,y)(\psi(y)+\epsilon\zeta(y))dy - \frac{\psi(x)+\epsilon\zeta(x)}{n(x)}\int\limits_{\mathcal{T}}\alpha(x,y)(\psi(y)+\epsilon\zeta(y))dy\right) \biggl{|}_{\epsilon = 0}\\
&= r\int\limits_{\mathcal{T}}m(x,y)\zeta(y)dy - \frac{1}{n(x)}\left(\psi(x)\int\limits_{\mathcal{T}}\alpha(x,y)\zeta(y)dy + \zeta(x)\int\limits_{\mathcal{T}}\alpha(x,y)\psi(y)dy\right)
\end{align*}
Using this in equation \eqref{spatial_covariance_zeta},
\cite{rogers_demographic_2012} (and later~\cite{rogers_modes_2015}) have shown that the contribution of demographic stochasticity can lead to inhibition of branching, and thus, while the population undergoes infinitely many branching events in the infinite population prediction, this does not happen for finite populations.
\myfig{0.8}{Media/2.1_dem_stoch_effects.png}{\textbf{Effect of population size on evolutionary branching}. Two different realizations of the system \eqref{Rogers_logistic_BD} with $n(x) = \exp(-x^2/\sigma_{K}^2)$. Simulation parameters are $\sigma_{K} = 1.9, \sigma_{\alpha} = 0.7, \sigma^2_{m} = 0.05$ for \textbf{top:} $K = 1000$ and \textbf{bottom:} $K = 10000$. Each point represents an individual. Note that the model on top remains monomorphic whereas the model on the bottom exhibits evolutionary branching, where an initially monomorphic population evolves to become dimorphic.}{fig_dem_stoch}
An alternative `moment-based' method that avoids moving to Fourier space has also been used to study this phenomenon of evolutionary branching and clustering in finite populations~\citep{wakano_evolutionary_2013,debarre_evolutionary_2016}. These studies use the equation we derived in section \ref{sec_fun_theorems_var} for the variance of the trait in the population and compute the conditions required for the variance to `explode' (Equation A.23 in~\cite{debarre_evolutionary_2016} is exactly equivalent to equation \eqref{nD_stochastic_Price_variance} for their choice of functional forms upon converting their change in variance to an infinitesimal rate of change \emph{i.e.} a derivative). The method itself is relatively straightforward in principle (complications arise if the particular models are complicated) and I therefore do not explore it further in this thesis, but the broad results of such moment-based approaches is in agreement with the predictions made from the spectral methods employed in~\cite{rogers_demographic_2012} and~\cite{rogers_modes_2015}.

It is left as an exercise for the reader to verify by the same steps that if we instead have the birth rate functional $b(x|\phi) = \lambda\int m(x,y)\phi(y)dy$ (with $m(x,y)$ as defined in \eqref{Rogers_logistic_BD}) and the death rate functional $d(x|\phi) = \phi(x)\left(\mu+(\lambda-\mu)\phi(x)/K\right)$, the infinite-population limit yields the famous Fisher-KPP equation with growth rate $r=\lambda-\mu$ and diffusion constant $D = \lambda \sigma_m^2/2$.
